---
title: "Mini Project 4"
author: "Jordan Moody, Marisa Youngblood, and Tara Marin"
date: "5/3/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(macleish)
library(tidyverse)
library(lubridate)
library(sp)
library(rgdal)
library(rgeos)
library(ggplot2)
```

###Problem Statement
In the wooded parcel of MacLeish Field Station, farmers and Smith students alike work to collect sap from trees in the area. However, climate change and warmer temperatures have led to a disruption in once reliable seasons. Given that MacLeish is an integral part of a local economy of farmers, producers, and distributors, it is essential to improve research and development for predicting the start and end date for the sugaring season in order to deliver the highest yields possible.

```{r, message = FALSE}
#Creating weather data tables
mac <- etl("macleish") %>%
  etl_update()
whately <- mac %>%
  tbl("whately")
orchard <- mac %>%
  tbl("orchard")

#Making tables of daily max and min temperatures
daily_whately <- whately %>%
  mutate(the_date = date(when)) %>%
  group_by(the_date) %>%
  summarize(N = n(), minTemp = min(temperature), maxTemp = max(temperature), avgTemp = mean(temperature)) %>%
  collect()
daily_orchard <- orchard %>%
  mutate(the_date = date(when)) %>%
  group_by(the_date) %>%
  summarize(N = n(), minTemp = min(temperature), maxTemp = max(temperature), avgTemp = mean(temperature)) %>%
  collect()

#Making tables of sap-able days
refined_whately <- daily_whately %>%
  filter(minTemp < 0, maxTemp > 0) %>%
  mutate(date = ymd(`the_date`))
refined_orchard <- daily_orchard %>%
  filter(minTemp < 0, maxTemp > 0) %>%
  mutate(date = ymd(`the_date`))
```

##Best times to sap ASIDE
I don't know how pretty I can make this graph, so it may not be something we have t include, but it shows that the best monthes to sap are November through March. There are good sapping days in October and early April, but not enough and not consistently enough for it to be logical to continue the sapping during those months. (generally two days in mid october and three or four throughout april)

```{r}
#orchard station graph
refined_orchard %>%
  ggplot(aes(x = date, y = avgTemp)) +
  geom_point() + geom_smooth() +
  ggtitle("Orchard Station Sapping Periods") +
  xlab("Date") + ylab("Average Daily Temperature") +
  theme_bw()
#whately station graph
refined_whately %>%
  ggplot(aes(x = date, y = avgTemp)) +
  geom_point() + geom_smooth() +
  ggtitle("Whately Station Sapping Periods") +
  xlab("Date") + ylab("Average Daily Temperature") +
  theme_bw()
```
```{r}
#isolating a specific sapping period
isolated_whately <- refined_whately %>%
  subset(date > "2014-10-19" & date < "2015-04-25")
isolated_whately %>%
  ggplot(aes(x = date, y = avgTemp)) +
  geom_point() +
  ggtitle("2014-2015 Sapping Period") +
  xlab("Date") + ylab("Average Daily Temperature") +
  theme_bw()
```

##Quantity of maple syrup

```{r}

```

##ASIDE 
I don't have the sap table up yet but the totals for each year are:
2013 = 675
2014 = 255 (very sporadic data for collection, most days had NA)
2015 = 349 (also fairly sporadic)
2016 = no data for this year
2017 = 259.5

averages:
2013 = 18.24
2014 = 6.38
2015 = 14.54
2017 = 17.38